---
title: "taller_than_wider"
output: html_document
date: "`r Sys.Date()`"
params:
    input: "TTW_publication_data.tsv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!requireNamespace("caret", quietly = TRUE))
  install.packages("caret")
if (!requireNamespace("data.table", quietly = TRUE))
  install.packages("data.table")
if (!requireNamespace("pracma", quietly = TRUE))
  install.packages("pracma")
if (!requireNamespace("pROC", quietly = TRUE))
  install.packages("pROC")
if (!requireNamespace("epiR", quietly = TRUE))
  install.packages("epiR")
if (!requireNamespace("cutpointr", quietly = TRUE))
  install.packages("cutpointr")

library(caret)
library(data.table)
library(pracma)
library(pROC)
library(epiR)
library(cutpointr)

```

```{r color, results='markup'}
print(params$input)
```
## Taller-Than-Wide Analysis

This markdown file replicates the analysis published in "Taller-than-wide sonographic feature poorly discriminates benign and malignant follicular and oncocytic thyroid neoplasms."

Abbreviations:
Follicular adenoma: FA
Follicular thyroid cancer: FTC 
Follicular variant papillary: FV-PCT
NIFTP: NIFT
Oncocytic adenoma: OA
Oncocytic thyroid cancer: OTC
Hyperplastic nodule: HN
Adenomatoid nodule: AN 
Papillary thyroid cancer: PTC

### Load and inspect data
```{r}
# Load data
dat3 <- fread(params$input, sep='\t')

# Summarize nodule dimensions
dimentions <- dat3[, c("H_AP", "W_Transverse", "L_Longitudinal")]
dims <- sort(as.numeric(dat3[1, c("H_AP", "W_Transverse", "L_Longitudinal")]))



summary <- data.frame(cancer_type = character(9), 
                      average_height = numeric(9),
                      average_width = numeric(9),
                      N_taller_than_wider = numeric(9),
                      N_total = numeric(9),
                      percent_taller_than_wider = numeric(9))

counter = 1

for (i in c("Follicular_adenoma", "Follicular_thyroid_cancer", "Follicular_variant_papillary",  "NIFTP",
            "Oncocytic_adenoma", "Oncocytic_thyroid_cancer", "Hyperplastic_nodule", "Adenomatoid_nodule", 
            "Papillary_thyroid_cancer")){
  
  subset <- dat3[dat3$Diagnosis == i,  ]  
  
  summary$cancer_type[counter] <- i
  summary$average_height[counter] <- mean(as.numeric(subset$H_AP))
  summary$average_width[counter] <- mean(as.numeric(subset$W_Transverse))
  summary$N_taller_than_wider[counter] <- sum(subset$H_AP > subset$W_Transverse)
  summary$N_total[counter] <- nrow(subset)
  summary$percent_taller_than_wider[counter] <- summary$N_taller_than_wider[counter]/summary$N_total[counter] * 100
  
  counter = counter + 1
  
}

print(summary)
```

### Calculating sphericity index

Sphericity index is defined as si <- nthroot(dims[1]^2/(dims[2]*dims[3]) ,3)
where dims is a data frame with three nodule dimensions.

```{r}

sphericity_index <- function(dims){ 
  dims <- sort(as.numeric(dims))
  si <- nthroot(dims[1]^2/(dims[2]*dims[3]) ,3)
  return(si)
}

# adding sphericity index
dimentions <- dat3[, c("H_AP", "W_Transverse", "L_Longitudinal")]
dat3$si <- apply(dimentions, MARGIN = 1, sphericity_index)

```


### Calculating spherical shape defined as a ratio of longest to shortest dimension

As proposed by Pappa et al., Thyroid Nodule Shape Independently Predicts Risk of Malignancy. J Clin Endocrinol Metab. 2022 Apr 19;107(7):1865–1870. doi: 10.1210/clinem/dgac246

```{r}

sphericity_ratio <- function(dims){ 
  
  dims <- sort(as.numeric(dims))
  sr <- dims[3]/dims[1]
  return(sr)
}

# adding sphericity index
dimentions <- dat3[, c("H_AP", "W_Transverse", "L_Longitudinal")]
dat3$sr <- apply(dimentions, MARGIN = 1, sphericity_ratio)

cor.test(dat3$si, dat3$sr, method = "spearman") 

```



## FA + OA vs. FTC + FV-PTC + OTC + NIFTP

```{r}
# Format data
dat_backup <- dat3

dat3$diag <- NA

dat3$diag[dat3$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma")] <- 'benign'

dat3$diag[dat3$Diagnosis %in% c("Follicular_thyroid_cancer", "Follicular_variant_papillary",  
                                "NIFTP", "Oncocytic_thyroid_cancer")] <- 'malignant'

dat3 <- dat3[!is.na(dat3$diag), ]

dat3$ttw <- NA
dat3$ttw[dat3$H_AP > dat3$W_Transverse] <- "taller_than_wide"
dat3$ttw[dat3$H_AP <= dat3$W_Transverse] <- "wider_than_tall"

# contingency table
ct <- table(dat3[, c("diag", "ttw")])
table(dat3[, c("diag", "ttw")])

#chi-squared test
chisq.test(dat3$diag, dat3$ttw) # 0.08; not significant

dat3$predicted <- NA
dat3$predicted[dat3$ttw == "wider_than_tall"] <- "benign"
dat3$predicted[dat3$ttw == "taller_than_wide"] <- "malignant"

# Create vectors
expected_value <- factor(dat3$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat3$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)

# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value

ct_epi <- as.table(matrix(c(ct["malignant", "taller_than_wide"], 
                            ct["benign", "taller_than_wide"],
                            ct["malignant", "wider_than_tall"],
                            ct["benign", "wider_than_tall"]), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)

# comparing sphericity index
#############################
print ("SPHERICITY INDEX")

#first testing if SI is normally distributed
hist(dat_backup$si) # not normal - left skewed
shapiro.test(dat_backup$si) # p-value = 4.158e-12, therefore will use non-parametric Wilcoxon test

wilcox.test(dat3$si[dat3$diag == "benign"], dat3$si[dat3$diag == "malignant"])


quantile(dat3$si[dat3$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat3$si[dat3$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat3$diag <- factor(dat3$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat3$diag,
                 predictor = dat3$si,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

optimal_cut <- cutpointr(x = dat3$si, class = dat3$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))

# CONFUSION MATRIX for SI
print ("Confusion matrix for SPHERICITY index")
# Create vectors
expected_value <- factor(dat3$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat3$si>optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)



################################
print ("SPHERICITY RATIO")

#first testing if SI is normally distributed
hist(dat_backup$sr) # not normal 
shapiro.test(dat_backup$sr) # p-value < 2.2e-16, therefore will use non-parametric Wilcoxon test

wilcox.test(dat3$sr[dat3$diag == "benign"], dat3$sr[dat3$diag == "malignant"])


quantile(dat3$sr[dat3$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat3$sr[dat3$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat3$diag <- factor(dat3$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat3$diag,
                 predictor = dat3$sr,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

optimal_cut <- cutpointr(x = dat3$sr, class = dat3$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))


# CONFUSION MATRIX for SR
print ("Confusion matrix for SPHERICITY RATIO")
# Create vectors
expected_value <- factor(dat3$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat3$sr<optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)


```


## TTW-1.2. FA + OA vs. FTC + FV-PTC + OTC + NIFTP

```{r}
# Format data
dat3_ttw1_2 <- dat_backup

dat3_ttw1_2$diag <- NA

dat3_ttw1_2$diag[dat3_ttw1_2$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma")] <- 'benign'

dat3_ttw1_2$diag[dat3_ttw1_2$Diagnosis %in% c("Follicular_thyroid_cancer", "Follicular_variant_papillary",  
                                "NIFTP", "Oncocytic_thyroid_cancer")] <- 'malignant'

dat3_ttw1_2 <- dat3_ttw1_2[!is.na(dat3_ttw1_2$diag), ]

dat3_ttw1_2$ttw <- NA
dat3_ttw1_2$ttw[dat3_ttw1_2$H_AP/dat3_ttw1_2$W_Transverse >= 1.2] <- "taller_than_wide"
dat3_ttw1_2$ttw[dat3_ttw1_2$H_AP/dat3_ttw1_2$W_Transverse < 1.2] <- "wider_than_tall"

# contingency table
ct <- table(dat3_ttw1_2[, c("diag", "ttw")])
table(dat3_ttw1_2[, c("diag", "ttw")])

#chi-squared test
chisq.test(dat3_ttw1_2$diag, dat3_ttw1_2$ttw)

dat3_ttw1_2$predicted <- NA
dat3_ttw1_2$predicted[dat3_ttw1_2$ttw == "wider_than_tall"] <- "benign"
dat3_ttw1_2$predicted[dat3_ttw1_2$ttw == "taller_than_wide"] <- "malignant"

# Create vectors
expected_value <- factor(dat3_ttw1_2$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat3_ttw1_2$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)

# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(ct["malignant", "taller_than_wide"], 
                            ct["benign", "taller_than_wide"],
                            ct["malignant", "wider_than_tall"],
                            ct["benign", "wider_than_tall"]), nrow = 2, byrow = TRUE))
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)

```

## FA vs. FTC incluing TTW 1.2 ratio

```{r}
dat5 <- dat3[dat3$Diagnosis %in% c("Follicular_adenoma", "Follicular_thyroid_cancer"), ]
table(dat5[, c("diag", "ttw")])
chisq.test(dat5$diag, dat5$ttw)

print ("SPHERICITY INDEX")
# comparing sphericity index
wilcox.test(dat5$si[dat5$diag == "benign"], dat5$si[dat5$diag == "malignant"])

quantile(dat5$si[dat5$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat5$si[dat5$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat5$diag <- factor(dat5$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat5$diag,predictor = dat5$si,levels = c('benign', 'malignant'))
roc
ci.auc(roc)


print ("SPHERICITY RATIO")

wilcox.test(dat5$sr[dat5$diag == "benign"], dat5$sr[dat5$diag == "malignant"])

quantile(dat5$sr[dat5$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat5$sr[dat5$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat5$diag <- factor(dat5$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat5$diag,
           predictor = dat5$sr,
           levels = c('benign', 'malignant'))
roc
ci.auc(roc)



# testing 1.2 TTW ratio
dat5_ttw1_2 <- dat5
dat5_ttw1_2$ttw <- NA
dat5_ttw1_2$ttw[dat5_ttw1_2$H_AP/dat5_ttw1_2$W_Transverse >= 1.2] <- "taller_than_wide"
dat5_ttw1_2$ttw[dat5_ttw1_2$H_AP/dat5_ttw1_2$W_Transverse < 1.2] <- "wider_than_tall"

table(dat5_ttw1_2[, c("diag", "ttw")])
chisq.test(dat5_ttw1_2$diag, dat5_ttw1_2$ttw)

```
## OA vs. OTC

```{r}
dat6 <- dat3[dat3$Diagnosis %in% c("Oncocytic_adenoma", "Oncocytic_thyroid_cancer"), ]
table(dat6[, c("diag", "ttw")])
chisq.test(dat6$diag, dat6$ttw)

print ("SPHERICITY INDEX")
# comparing sphericity index
wilcox.test(dat6$si[dat6$diag == "benign"], dat6$si[dat6$diag == "malignant"])

quantile(dat6$si[dat6$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat6$si[dat6$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat6$diag <- factor(dat6$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat6$diag,
                 predictor = dat6$si,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

print ("SPHERICITY RATIO")
wilcox.test(dat6$sr[dat6$diag == "benign"], dat6$sr[dat6$diag == "malignant"])

quantile(dat6$sr[dat6$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat6$sr[dat6$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat6$diag <- factor(dat6$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat6$diag,
                 predictor = dat6$sr,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)


# testing 1.2 TTW ratio
dat6_ttw1_2 <- dat6
dat6_ttw1_2$ttw <- NA
dat6_ttw1_2$ttw[dat6_ttw1_2$H_AP/dat6_ttw1_2$W_Transverse >= 1.2] <- "taller_than_wide"
dat6_ttw1_2$ttw[dat6_ttw1_2$H_AP/dat6_ttw1_2$W_Transverse < 1.2] <- "wider_than_tall"

table(dat6_ttw1_2[, c("diag", "ttw")])
chisq.test(dat6_ttw1_2$diag, dat6_ttw1_2$ttw)

```



## PTC vs. AN + HN

```{r}
# Format data
dat8 <- dat_backup[dat_backup$Diagnosis %in% c("Adenomatoid_nodule", "Hyperplastic_nodule", "Papillary_thyroid_cancer"), ]


dat8$diag <- NA

dat8$diag[dat8$Diagnosis %in% c("Adenomatoid_nodule", "Hyperplastic_nodule")] <- 'benign'

dat8$diag[dat8$Diagnosis %in% c("Papillary_thyroid_cancer")] <- 'malignant'

dat8 <- dat8[!is.na(dat8$diag), ]

dat8$ttw <- NA
dat8$ttw[dat8$H_AP > dat8$W_Transverse] <- "taller_than_wide"
dat8$ttw[dat8$H_AP <= dat8$W_Transverse] <- "wider_than_tall"

# contingency table
ct <- table(dat8[, c("diag", "ttw")])
table(dat8[, c("diag", "ttw")])

chisq.test(dat8$diag, dat8$ttw) 

# confusion matrix
dat8$predicted <- NA
dat8$predicted[dat8$ttw == "wider_than_tall"] <- "benign"
dat8$predicted[dat8$ttw == "taller_than_wide"] <- "malignant"

# Create vectors
expected_value <- factor(dat8$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat8$predicted, levels = c("malignant", "benign"))


# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)

# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(ct["malignant", "taller_than_wide"], 
                            ct["benign", "taller_than_wide"],
                            ct["malignant", "wider_than_tall"],
                            ct["benign", "wider_than_tall"]), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 2, conf.level = 0.95)



###########################
print ("SPHERICITY INDEX")
# comparing sphericity index
wilcox.test(dat8$si[dat8$diag == "benign"], dat8$si[dat8$diag == "malignant"])

quantile(dat8$si[dat8$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat8$si[dat8$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))


# calculating AUROC
dat8$diag <- factor(dat8$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat8$diag,
                 predictor = dat8$si,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

optimal_cut <- cutpointr(x = dat8$si, class = dat8$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))

# CONFUSION MATRIX for SI
print ("Confusion matrix for SPHERICITY index")
# Create vectors
expected_value <- factor(dat8$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat8$si>optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)



################################
print ("SPHERICITY RATIO")

wilcox.test(dat8$sr[dat8$diag == "benign"], dat8$sr[dat8$diag == "malignant"])


quantile(dat8$sr[dat8$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat8$sr[dat8$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat8$diag <- factor(dat8$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat8$diag,
                 predictor = dat8$sr,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

optimal_cut <- cutpointr(x = dat8$sr, class = dat8$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))


# CONFUSION MATRIX for SR
print ("Confusion matrix for SPHERICITY RATIO")
# Create vectors
expected_value <- factor(dat8$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat8$sr<optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)


#######################
# testing 1.2 TTW ratio
dat8_ttw1_2 <- dat8
dat8_ttw1_2$ttw <- NA
dat8_ttw1_2$ttw[dat8_ttw1_2$H_AP/dat8_ttw1_2$W_Transverse >= 1.2] <- "taller_than_wide"
dat8_ttw1_2$ttw[dat8_ttw1_2$H_AP/dat8_ttw1_2$W_Transverse < 1.2] <- "wider_than_tall"

table(dat8_ttw1_2[, c("diag", "ttw")])
chisq.test(dat8_ttw1_2$diag, dat8_ttw1_2$ttw)



```

## FA + OA vs. FTC + OTC

```{r}
# Format data
dat9 <- dat_backup
dat9 <- dat9[dat9$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma", "Follicular_thyroid_cancer", "Oncocytic_thyroid_cancer"), ]
dat9$diag <- NA

dat9$diag[dat9$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma")] <- 'benign'
dat9$diag[dat9$Diagnosis %in% c("Follicular_thyroid_cancer", "Oncocytic_thyroid_cancer")] <- 'malignant'
dat9 <- dat9[!is.na(dat9$diag), ]

dat9$ttw <- NA

dat9$ttw[dat9$H_AP > dat9$W_Transverse] <- "taller_than_wide"
dat9$ttw[dat9$H_AP <= dat9$W_Transverse] <- "wider_than_tall"

table(dat9[, c("diag", "ttw")])
chisq.test(dat9$diag, dat9$ttw) 

# contingency table
ct <- table(dat9[, c("diag", "ttw")])
table(dat9[, c("diag", "ttw")])


dat9$predicted <- NA
dat9$predicted[dat9$ttw == "wider_than_tall"] <- "benign"
dat9$predicted[dat9$ttw == "taller_than_wide"] <- "malignant"

# Create vectors
expected_value <- factor(dat9$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat9$predicted, levels = c("malignant", "benign"))


# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)

# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(ct["malignant", "taller_than_wide"], 
                            ct["benign", "taller_than_wide"],
                            ct["malignant", "wider_than_tall"],
                            ct["benign", "wider_than_tall"]), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 2, conf.level = 0.95)


###########################
print ("SPHERICITY INDEX")
# comparing sphericity index
wilcox.test(dat9$si[dat9$diag == "benign"], dat9$si[dat9$diag == "malignant"])

quantile(dat9$si[dat9$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat9$si[dat9$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))


# calculating AUROC
dat9$diag <- factor(dat9$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat9$diag,
                 predictor = dat9$si,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)


optimal_cut <- cutpointr(x = dat9$si, class = dat9$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))

# CONFUSION MATRIX for SI
print ("Confusion matrix for SPHERICITY index")
# Create vectors
expected_value <- factor(dat9$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat9$si>optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)



################################
print ("SPHERICITY RATIO")

wilcox.test(dat9$sr[dat9$diag == "benign"], dat9$sr[dat9$diag == "malignant"])


quantile(dat9$sr[dat9$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat9$sr[dat9$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat9$diag <- factor(dat9$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat9$diag,
                 predictor = dat9$sr,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

optimal_cut <- cutpointr(x = dat9$sr, class = dat9$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))


# CONFUSION MATRIX for SR
print ("Confusion matrix for SPHERICITY RATIO")
# Create vectors
expected_value <- factor(dat9$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat9$sr<optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)

##########################
# testing 1.2 TTW ratio
dat9_ttw1_2 <- dat9
dat9_ttw1_2$ttw <- NA
dat9_ttw1_2$ttw[dat9_ttw1_2$H_AP/dat9_ttw1_2$W_Transverse >= 1.2] <- "taller_than_wide"
dat9_ttw1_2$ttw[dat9_ttw1_2$H_AP/dat9_ttw1_2$W_Transverse < 1.2] <- "wider_than_tall"

table(dat9_ttw1_2[, c("diag", "ttw")])
chisq.test(dat9_ttw1_2$diag, dat9_ttw1_2$ttw)


```


## HN + AN vs. FA + OA + FTC + NIFTP + FVP + OTC + PTC (non_neoplastic vs. neoplastic)

```{r}
# Format data
dat10 <- dat_backup
table(dat10$Diagnosis)

dat10$diag <- NA

dat10$diag[dat10$Diagnosis %in% c("Hyperplastic_nodule", "Adenomatoid_nodule")] <- 'non_neoplastic'
dat10$diag[dat10$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma", "Follicular_thyroid_cancer", 
                                  "Follicular_variant_papillary",  
                                  "NIFTP", "Oncocytic_thyroid_cancer", "Papillary_thyroid_cancer")] <- 'neoplastic'
dat10 <- dat10[!is.na(dat10$diag), ]

dat10$ttw <- NA

dat10$ttw[dat10$H_AP > dat10$W_Transverse] <- "taller_than_wide"
dat10$ttw[dat10$H_AP <= dat10$W_Transverse] <- "wider_than_tall"

table(dat10[, c("diag", "ttw")])
chisq.test(dat10$diag, dat10$ttw)

# contingency table
ct <- table(dat10[, c("diag", "ttw")])
table(dat10[, c("diag", "ttw")])

dat10$predicted <- NA
dat10$predicted[dat10$ttw == "wider_than_tall"] <- "benign"
dat10$predicted[dat10$ttw == "taller_than_wide"] <- "malignant"

# Convert to proper comparison
dat10$diag <- ifelse(dat10$diag == "neoplastic",
                     "malignant",
                     dat10$diag)
dat10$diag <- ifelse(dat10$diag == "non_neoplastic",
                     "benign",
                     dat10$diag)

# Create vectors
expected_value <- factor(dat10$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat10$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)

# using epiR to calculate exact binomial confidence limits for test sensitivity, specificity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(ct["neoplastic", "taller_than_wide"], 
                            ct["non_neoplastic", "taller_than_wide"],
                            ct["neoplastic", "wider_than_tall"],
                            ct["non_neoplastic", "wider_than_tall"]), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)

print ("SPHERICITY INDEX")
# comparing sphericity index
wilcox.test(dat10$si[dat10$diag == "benign"], dat10$si[dat10$diag == "malignant"])

quantile(dat10$si[dat10$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat10$si[dat10$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))


# calculating AUROC
dat10$diag <- factor(dat10$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat10$diag,
                 predictor = dat10$si,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

optimal_cut <- cutpointr(x = dat9$si, class = dat9$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))

# CONFUSION MATRIX for SI
print ("Confusion matrix for SPHERICITY index")
# Create vectors
expected_value <- factor(dat10$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat10$si>optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)



################################
print ("SPHERICITY RATIO")

wilcox.test(dat10$sr[dat10$diag == "benign"], dat10$sr[dat10$diag == "malignant"])


quantile(dat10$sr[dat10$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat10$sr[dat10$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat10$diag <- factor(dat10$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat10$diag,
                 predictor = dat10$sr,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

optimal_cut <- cutpointr(x = dat10$sr, class = dat10$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))


# CONFUSION MATRIX for SR
print ("Confusion matrix for SPHERICITY RATIO")
# Create vectors
expected_value <- factor(dat10$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat10$sr<optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)


#############################
# testing 1.2 TTW ratio
dat10_ttw1_2 <- dat10
dat10_ttw1_2$ttw <- NA
dat10_ttw1_2$ttw[dat10_ttw1_2$H_AP/dat10_ttw1_2$W_Transverse >= 1.2] <- "taller_than_wide"
dat10_ttw1_2$ttw[dat10_ttw1_2$H_AP/dat10_ttw1_2$W_Transverse < 1.2] <- "wider_than_tall"

table(dat10_ttw1_2[, c("diag", "ttw")])
chisq.test(dat10_ttw1_2$diag, dat10_ttw1_2$ttw)





```


## HN + AN + FA + OA vs. FTC + NIFTP + FVP + OTC + PTC (benign vs. malignant)

```{r}
# Format data
dat11 <- dat_backup
table(dat11$Diagnosis)

dat11$diag <- NA

dat11$diag[dat11$Diagnosis %in% c("Hyperplastic_nodule", "Adenomatoid_nodule", "Follicular_adenoma", "Oncocytic_adenoma")] <- 'benign'
dat11$diag[dat11$Diagnosis %in% c("Follicular_thyroid_cancer", "Follicular_variant_papillary",  
                                  "NIFTP", "Oncocytic_thyroid_cancer", "Papillary_thyroid_cancer")] <- 'malignant'
dat11 <- dat11[!is.na(dat11$diag), ]

dat11$ttw <- NA

dat11$ttw[dat11$H_AP > dat11$W_Transverse] <- "taller_than_wide"
dat11$ttw[dat11$H_AP <= dat11$W_Transverse] <- "wider_than_tall"

table(dat11[, c("diag", "ttw")])
chisq.test(dat11$diag, dat11$ttw)

# contingency table
ct <- table(dat11[, c("diag", "ttw")])
table(dat11[, c("diag", "ttw")])

dat11$predicted <- NA
dat11$predicted[dat11$ttw == "wider_than_tall"] <- "benign"
dat11$predicted[dat11$ttw == "taller_than_wide"] <- "malignant"

# Create vectors
expected_value <- factor(dat11$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat11$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)

# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(ct["malignant", "taller_than_wide"], 
                            ct["benign", "taller_than_wide"],
                            ct["malignant", "wider_than_tall"],
                            ct["benign", "wider_than_tall"]), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)

print ("SPHERICITY INDEX")
# comparing sphericity index
wilcox.test(dat11$si[dat11$diag == "benign"], dat11$si[dat11$diag == "malignant"])

quantile(dat11$si[dat11$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat11$si[dat11$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))


# calculating AUROC
dat11$diag <- factor(dat11$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat11$diag,
                 predictor = dat11$si,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

optimal_cut <- cutpointr(x = dat11$si, class = dat11$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))

# CONFUSION MATRIX for SI
print ("Confusion matrix for SPHERICITY index")
# Create vectors
expected_value <- factor(dat11$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat11$si>optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)



################################
print ("SPHERICITY RATIO")

wilcox.test(dat11$sr[dat11$diag == "benign"], dat11$sr[dat11$diag == "malignant"])


quantile(dat11$sr[dat11$diag == "benign"], probs = c(0,0.25,0.5,0.75,1))
quantile(dat11$sr[dat11$diag == "malignant"], probs = c(0,0.25,0.5,0.75,1))

# calculating AUROC
dat11$diag <- factor(dat11$diag, levels = c("benign", "malignant"))

roc <- pROC::roc(response = dat11$diag,
                 predictor = dat11$sr,
                 levels = c('benign', 'malignant'))
roc
ci.auc(roc)

optimal_cut <- cutpointr(x = dat11$sr, class = dat11$diag, method = maximize_metric, metric = youden) 
print(paste0("Optimal cutpoint: ", optimal_cut$optimal_cutpoint))


# CONFUSION MATRIX for SR
print ("Confusion matrix for SPHERICITY RATIO")
# Create vectors
expected_value <- factor(dat11$diag, levels = c("malignant", "benign"))
predicted_value <- factor(ifelse(dat11$sr<optimal_cut$optimal_cutpoint, "malignant", "benign"), levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
# using epiR to calculate exact binomial confidence limits for test sensitivity, specif icity, and positive and negative predictive value
ct_epi <- as.table(matrix(c(sum(expected_value == "malignant" & predicted_value == "malignant"), 
                            sum(expected_value == "benign" & predicted_value == "malignant"),
                            sum(expected_value == "malignant" & predicted_value == "benign"),
                            sum(expected_value == "benign" & predicted_value == "benign")), nrow = 2, byrow = TRUE))
ct_epi
epi.tests(ct_epi, method = "exact", digits = 3, conf.level = 0.95)


###########################
# testing 1.2 TTW ratio
dat11_ttw1_2 <- dat11
dat11_ttw1_2$ttw <- NA
dat11_ttw1_2$ttw[dat11_ttw1_2$H_AP/dat11_ttw1_2$W_Transverse >= 1.2] <- "taller_than_wide"
dat11_ttw1_2$ttw[dat11_ttw1_2$H_AP/dat11_ttw1_2$W_Transverse < 1.2] <- "wider_than_tall"

table(dat11_ttw1_2[, c("diag", "ttw")])
chisq.test(dat11_ttw1_2$diag, dat11_ttw1_2$ttw)





```