---
title: "taller_than_wider"
output: html_document
date: "`r Sys.Date()`"
params:
    input: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("readxl", quietly = TRUE))
  install.packages("readxl")
if (!requireNamespace("caret", quietly = TRUE))
  install.packages("caret")
if (!requireNamespace("data.table", quietly = TRUE))
  install.packages("pracma")
if (!requireNamespace("pracma", quietly = TRUE))
  install.packages("pROC")
if (!requireNamespace("yardstick", quietly = TRUE))
  install.packages("yardstick")

library(readxl)
library(caret)
library(data.table)
library(pracma)
library(pROC)
library(yardstick)
```

```{r color, results='markup'}
print(params$input)
```
## Taller-Than-Wide Analysis

This markdown file replicates the analysis published in "Taller-than-wide sonographic feature poorly discriminates benign and malignant follicular and oncocytic thyroid neoplasms."

Abbreviations:
Follicular adenoma: FA
Follicular thyroid cancer: FTC 
Follicular variant papillary: FVP
NIFTP: NIFT
Oncocytic adenoma: OA
Oncocytic thyroid cancer: OTC
Hyperplastic nodule: HN
Adenomatoid nodule: AN 
Papillary thyroid cancer: PTC

### Load and inspect data
```{r}
# Load data
dat3 <- fread(params$input, sep='\t')

# Summarize nodule dimensions
dimentions <- dat3[, c("H_AP", "W_Transverse", "L_Longitudinal")]
dims <- sort(as.numeric(dat3[1, c("H_AP", "W_Transverse", "L_Longitudinal")]))



summary <- data.frame(cancer_type = character(9), 
                      average_height = numeric(9),
                      average_width = numeric(9),
                      N_taller_than_wider = numeric(9),
                      N_total = numeric(9),
                      percent_taller_than_wider = numeric(9))

counter = 1

for (i in c("Follicular_adenoma", "Follicular_thyroid_cancer", "Follicular_variant_papillary",  "NIFTP",
            "Oncocytic_adenoma", "Oncocytic_thyroid_cancer", "Hyperplastic_nodule", "Adenomatoid_nodule", 
            "Papillary_thyroid_cancer")){
  
  subset <- dat3[dat3$Diagnosis == i,  ]  
  
  summary$cancer_type[counter] <- i
  summary$average_height[counter] <- mean(as.numeric(subset$H_AP))
  summary$average_width[counter] <- mean(as.numeric(subset$W_Transverse))
  summary$N_taller_than_wider[counter] <- sum(subset$H_AP > subset$W_Transverse)
  summary$N_total[counter] <- nrow(subset)
  summary$percent_taller_than_wider[counter] <- summary$N_taller_than_wider[counter]/summary$N_total[counter] * 100
  
  counter = counter + 1
  
}

print(summary)
```

## FA + OA vs. FTC + FV-PTC + OTC + NIFTP

```{r}
# Format data
dat_backup <- dat3

dat3$diag <- NA

dat3$diag[dat3$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma")] <- 'benign'

dat3$diag[dat3$Diagnosis %in% c("Follicular_thyroid_cancer", "Follicular_variant_papillary",  
                                "NIFTP", "Oncocytic_thyroid_cancer")] <- 'malignant'

dat3 <- dat3[!is.na(dat3$diag), ]

dat3$ttw <- NA
dat3$ttw[dat3$H_AP > dat3$W_Transverse] <- "taller_than_wide"
dat3$ttw[dat3$H_AP <= dat3$W_Transverse] <- "wider_than_tall"

table(dat3[, c("diag", "ttw")])
chisq.test(dat3$diag, dat3$ttw) # NOT SIGNIFICANT!!!!

dat3$predicted <- NA
dat3$predicted[dat3$ttw == "wider_than_tall"] <- "benign"
dat3$predicted[dat3$ttw == "taller_than_wide"] <- "malignant"

# Create vectors
expected_value <- factor(dat3$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat3$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
```

## FA + OA vs. FTC + FV-PTC + OTC

```{r}
# Format data
dat4 <- dat3[dat3$Diagnosis != "NIFTP", ]

dat4$diag <- NA

dat4$diag[dat4$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma")] <- 'benign'
dat4$diag[dat4$Diagnosis %in% c("Follicular_thyroid_cancer", "Follicular_variant_papillary",  
                                "NIFTP", "Oncocytic_thyroid_cancer")] <- 'malignant'
dat4 <- dat4[!is.na(dat4$diag), ]


dat4$ttw <- NA

dat4$ttw[dat4$H_AP > dat4$W_Transverse] <- "taller_than_wide"
dat4$ttw[dat4$H_AP <= dat4$W_Transverse] <- "wider_than_tall"

table(dat4[, c("diag", "ttw")])
chisq.test(dat4$diag, dat4$ttw) 

dat4$predicted <- NA
dat4$predicted[dat4$ttw == "wider_than_tall"] <- "benign"
dat4$predicted[dat4$ttw == "taller_than_wide"] <- "malignant"

# Create vectors
expected_value <- factor(dat4$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat4$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
```

## FA vs. FTC

```{r}
dat5 <- dat4[dat4$Diagnosis %in% c("Follicular_adenoma", "Follicular_thyroid_cancer"), ]
table(dat5[, c("diag", "ttw")])
chisq.test(dat5$diag, dat5$ttw)
```

## OA vs. OTC

```{r}
dat6 <- dat4[dat4$Diagnosis %in% c("Oncocytic_adenoma", "Oncocytic_thyroid_cancer"), ]
table(dat6[, c("diag", "ttw")])
chisq.test(dat6$diag, dat6$ttw)
```

## FA + OA + AN + HN vs. FTC + FVP + NIFTP + OCT + PTC

```{r}
# Foramt data
dat7 <- dat_backup
table(dat7$Diagnosis)

dat7$diag <- NA

dat7$diag[dat7$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma", 
                                "Adenomatoid_nodule", "Hyperplastic_nodule")] <- 'benign'
dat7$diag[dat7$Diagnosis %in% c("Follicular_thyroid_cancer", "Follicular_variant_papillary",  
                                "NIFTP", "Oncocytic_thyroid_cancer", "Papillary_thyroid_cancer")] <- 'malignant'
dat7 <- dat7[!is.na(dat7$diag), ]

dat7$ttw <- NA

dat7$ttw[dat7$H_AP > dat7$W_Transverse] <- "taller_than_wide"
dat7$ttw[dat7$H_AP <= dat7$W_Transverse] <- "wider_than_tall"


table(dat7[, c("diag", "ttw")])
chisq.test(dat7$diag, dat7$ttw)

dat7$predicted <- NA
dat7$predicted[dat7$ttw == "wider_than_tall"] <- "benign"
dat7$predicted[dat7$ttw == "taller_than_wide"] <- "malignant"

# Crete vectors
expected_value <- factor(dat7$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat7$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
```

## PTC vs. AN + HN

```{r}
# Format data
dat8 <- dat7[dat7$Diagnosis %in% c("Adenomatoid_nodule", "Hyperplastic_nodule", "Papillary_thyroid_cancer"), ]

table(dat8[, c("diag", "ttw")])
chisq.test(dat8$diag, dat8$ttw) 

# Create vectors
expected_value <- factor(dat8$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat8$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
```

## FA + OA vs. FTC + OTC

```{r}
# Format data
dat9 <- dat_backup
dat9 <- dat9[dat9$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma", "Follicular_thyroid_cancer", "Oncocytic_thyroid_cancer"), ]
dat9$diag <- NA

dat9$diag[dat9$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma")] <- 'benign'
dat9$diag[dat9$Diagnosis %in% c("Follicular_thyroid_cancer", "Oncocytic_thyroid_cancer")] <- 'malignant'
dat9 <- dat9[!is.na(dat9$diag), ]

dat9$ttw <- NA

dat9$ttw[dat9$H_AP > dat9$W_Transverse] <- "taller_than_wide"
dat9$ttw[dat9$H_AP <= dat9$W_Transverse] <- "wider_than_tall"

table(dat9[, c("diag", "ttw")])
chisq.test(dat9$diag, dat9$ttw) 

dat9$predicted <- NA
dat9$predicted[dat9$ttw == "wider_than_tall"] <- "benign"
dat9$predicted[dat9$ttw == "taller_than_wide"] <- "malignant"

# Create vectors
expected_value <- factor(dat9$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat9$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
```

## PTC vs. FA + OA + FTC + OTC + NIFTP + FVPTC

```{r}
# Format data
dat10 <- dat_backup
dat10 <- dat10[dat10$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma", "Follicular_thyroid_cancer", "Oncocytic_thyroid_cancer",
                                      "Follicular_variant_papillary", "NIFTP", "Papillary_thyroid_cancer"), ]
dat10$diag <- NA

dat10$diag[dat10$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma", "Follicular_thyroid_cancer", "Oncocytic_thyroid_cancer",
                                  "Follicular_variant_papillary", "NIFTP")] <- 'nonPTC_neoplasms'
dat10$diag[dat10$Diagnosis %in% c("Papillary_thyroid_cancer")] <- 'PTC'
dat10 <- dat10[!is.na(dat10$diag), ]

dat10$ttw <- NA

dat10$ttw[dat10$H_AP > dat10$W_Transverse] <- "taller_than_wide"
dat10$ttw[dat10$H_AP <= dat10$W_Transverse] <- "wider_than_tall"

table(dat10[, c("diag", "ttw")])
chisq.test(dat10$diag, dat10$ttw) 

dat10$predicted <- NA
dat10$predicted[dat10$ttw == "wider_than_tall"] <- "nonPTC_neoplasms"
dat10$predicted[dat10$ttw == "taller_than_wide"] <- "PTC"

# Create vectors
expected_value <- factor(dat10$diag, levels = c("PTC", "nonPTC_neoplasms"))
predicted_value <- factor(dat10$predicted, levels = c("PTC", "nonPTC_neoplasms"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
```

## HN + AN vs. FA + OA + FTC + NIFTP + FVP + OTC

```{r}
# Format data
dat10 <- dat_backup
table(dat10$Diagnosis)

dat10$diag <- NA

dat10$diag[dat10$Diagnosis %in% c("Hyperplastic_nodule", "Adenomatoid_nodule")] <- 'non_neoplastic'
dat10$diag[dat10$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma", "Follicular_thyroid_cancer", 
                                  "Follicular_variant_papillary",  
                                  "NIFTP", "Oncocytic_thyroid_cancer")] <- 'neoplastic'
dat10 <- dat10[!is.na(dat10$diag), ]

dat10$ttw <- NA

dat10$ttw[dat10$H_AP > dat10$W_Transverse] <- "taller_than_wide"
dat10$ttw[dat10$H_AP <= dat10$W_Transverse] <- "wider_than_tall"

table(dat10[, c("diag", "ttw")])
chisq.test(dat10$diag, dat10$ttw) # NOT SIGNIFICANT!!!!

dat10$predicted <- NA
dat10$predicted[dat10$ttw == "wider_than_tall"] <- "benign"
dat10$predicted[dat10$ttw == "taller_than_wide"] <- "malignant"

# Convert to proper comparison
dat10$diag <- ifelse(dat10$diag == "neoplastic",
                     "malignant",
                     dat10$diag)
dat10$diag <- ifelse(dat10$diag == "non_neoplastic",
                     "benign",
                     dat10$diag)

# Create vectors
expected_value <- factor(dat10$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat10$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
```

## PTC vs. FV-PTC

```{r}
# Format data
dat10 <- dat_backup
table(dat10$Diagnosis)

dat10$diag <- NA

dat10$diag[dat10$Diagnosis %in% c("Papillary_thyroid_cancer")] <- 'PTC'
dat10$diag[dat10$Diagnosis %in% c("Follicular_variant_papillary")] <- 'FV-PTC'
dat10 <- dat10[!is.na(dat10$diag), ]

dat10$ttw <- NA

dat10$ttw[dat10$H_AP > dat10$W_Transverse] <- "taller_than_wide"
dat10$ttw[dat10$H_AP <= dat10$W_Transverse] <- "wider_than_tall"

table(dat10[, c("diag", "ttw")])
chisq.test(dat10$diag, dat10$ttw) # NOT SIGNIFICANT!!!!

dat10$predicted <- NA
dat10$predicted[dat10$ttw == "wider_than_tall"] <- "benign"
dat10$predicted[dat10$ttw == "taller_than_wide"] <- "malignant"

# Convert to proper comparison
dat10$diag <- ifelse(dat10$diag == "PTC",
                     "malignant",
                     dat10$diag)
dat10$diag <- ifelse(dat10$diag == "FV-PTC",
                     "benign",
                     dat10$diag)

# Create vectors
expected_value <- factor(dat10$diag, levels = c("malignant", "benign"))
predicted_value <- factor(dat10$predicted, levels = c("malignant", "benign"))

# Create confusion matrix
confusionMatrix(data=predicted_value, reference = expected_value)
```

## Regression to investigate significance of nodule size

```{r}
# Format data
dat11 <- dat_backup

dat11$diag <- NA

dat11$diag[dat11$Diagnosis %in% c("Follicular_adenoma", "Oncocytic_adenoma")] <- 'benign'
dat11$diag[dat11$Diagnosis %in% c("Follicular_thyroid_cancer", "Follicular_variant_papillary",  
                                  "NIFTP", "Oncocytic_thyroid_cancer")] <- 'malignant'
dat11 <- dat11[!is.na(dat11$diag), ]

dat11$ttw <- NA

dat11$ttw[dat11$H_AP > dat11$W_Transverse] <- 1
dat11$ttw[dat11$H_AP <= dat11$W_Transverse] <- 0

# Max dimension
dat11$maxNoduleSize <- apply(dat11[, c("H_AP", "W_Transverse", "L_Longitudinal")], max, MARGIN=1)
dat11$diag <- as.numeric(as.factor(dat11$Diagnosis))

# Regression
model <- glm(formula = ttw ~ diag + maxNoduleSize + diag * maxNoduleSize, 
    family = "binomial", data = dat11)

summary(model)
```